{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-09-01 thesis/education.tex
final pre-send-to-committee revisions to education chapater

The (unpatched versions of) the first four tests are available at
\url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
The fifth test is available at
\url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.

@@

\begin{table}[h]
	\begin{center}
		% ugh this looks so stupid
		% but better than latex's output looking stupid
		\begin{tabular}{r|crcrc|r}
			& \multicolumn{5}{c|}{\bf Used Landslide?} & \\
			\bf Semester & & \bf yes & & \bf no & & \bf Total \\
			\hline
			CMU S'15	& &   8	& &  18	& &  26 \\
			CMU F'15	& &  22	& &   8	& &  30 \\
			CMU S'16	& &  18	& &  16	& &  34 \\
			CMU F'16	& &  12	& &   6	& &  18 \\
			CMU S'17	& &  19	& &  13	& &  32 \\
			CMU F'17	& &  14	& &   5	& &  19 \\
			CMU S'18	& &  10	& &   8	& &  18 \\
			\hline
			U. Chicago F'17	& &   4	& &  17	& &  21	\\ % 21 taken from number of submitted 'threads', idk abt userprog
			% TODO when you adjust this, adjust the % of participance down in the discussion para
			PSU S'18	& &  38	& &   ?	& &   ?	\\
			\hline
			\bf CMU total	& & 103	& &  74	& & 177 \\
			\bf Non-CMU total
					& &  42	& & 17+?& &  21+? \\
			\bf Total	& & 145 & & 91+?& & 191+? \\
		\end{tabular}
	\end{center}
	\caption{Participation rates across semesters,
	among students who submitted thread libraries (CMU, PSU) or kernels (U. Chicago).}
	\label{tab:photo-of-ze-studence}
\end{table}

@@

In the recruiting lectures for every semester after the first,
I included tallies of Landslide's bug-finding achievements to date
as an extra advertisement to tempt students.
These included each of the measures I counted by hand as described above,
as well as aggregate totals of how many groups participated,
how many groups found bugs,
how many found at least one concurrency bug,
and how many were able to fix and subsequently verify any or all thereof.
Among the concurrency bugs, a wide variety of types were found:
deadlocks, use-after-frees, segfaults, infinite loops, assertion failure, and unit test failure.
Table~\ref{tab:cmubugs} shows these statistics for each semester.
% may also be an overestimate if they only hacked around the bug, but that's less likely
Considering the method for deciding whether students fixed each bug, described above,
note that the tallies thereof may be undercounting
on account of the possibility some students may have skipped the verification step despite truly fixing them.

\begin{table}[t]
	% TODO
	\begin{center}
		\small
	\begin{tabular}{r||r|r|r|r|r|r|r||r}
			& \bf S'15 & \bf F'15 & \bf S'16 & \bf F'16 & \bf S'17 & \bf F'17 & \bf S'18 & \bf Total \\
			\hline
			%			S15	F15	S16	F16	S17	F17	S18
			\bf Participating groups& 8	& 22	& 18	& 12	& 19	& 14	& 10	& 103	\\
			% nb not counting FPs
			Groups w/any bugs found	& 5	& 21	& 12	& 7	& 14	& 10	& 6	& 75	\\
			Groups w/any bugs fixed	& 4	& 18	& 9	& 6	& 9	& 7	& 5	& 58	\\
			Groups w/all (>0) bugs fixed
						& 2	& 11	& 3	& 3	& 7	& 5	& 4	& 35	\\
			\hline
			% the 2 for f15 were pretty messy; i had to go back and get em by hand by grepping for NOTE
			% but this semester also had a LOT of seemingly true race NOTE mx lock inf loop,
			% so i counted only the det ones, of which there were 2 (both also in "fixed" category)
			% otherwise 1 race recursive mx and 6 race mutex loops and 1 race cvar loop were left as races.

@@

This table also shows the number of false positives, as self-reported by Landslide
as described in \sect{\ref{sec:education-pebbles-instrumentation}}.
% nah where actually doing this, where making it hapen.
%which I started counting since S'16.
Please note that this approach %to counting false positives
is limited by Landslide's ability to classify them as ``suspicious'', even if not definitely bugs.
These represent technical obstacles that either
prevented Landslide from being able to analyze the synchronization in play
(e.g., recursive {\tt mutex\_lock()} invocations)
or were deemed too pedagogically important to allow the student to proceed without fixing
(e.g., busy spin-wait synchronization loops)%
\footnote{
Landslide was also configured to issue a ``bug report''
if {\tt MAGIC\_BREAK}, a Simics debugging trap, was ever invoked,
with specific instructions to remove it;
I consider these closer to compilation errors than bug reports and so did not count them
even among false positives.
}.
Other false positives arising from bugs in Landslide itself
required more individual effort to confirm;
to classify these I relied on students to help report them
during the Landslide clinics and survey,
and I report on them in \sect{\ref{sec:education-eval-survey-falsepositives}}.
Nevertheless,
I am aware of no case in which Landslide would output a false positive {\em nondeterministic} bug%
% this is really shaky
% not too shaky for me to feel uncomfortable publishing it
% but too shaky for me to feel comfortable leaving it unsaid too...
\footnote{I did observe Landslide's ``suspicious loop'' message
for mutex or condition variable internal synchronization
on subsequent interleavings,
but at this point the line between style warning and true bug blurs,
so I count these as true concurrency bugs for simplicity's sake.}
% i *think*.......
all unexpected false positives encountered to date were output on the first interleaving tested.
Though unscientific,
this provides some reassurance that Table~\ref{tab:cmubugs}'s count of concurrency bugs is accurate,
even if the deterministic tally may overcount slightly.

@@

In cases where Landslide emitted multiple preemption traces for the same bug,
whether manifested the same way through different combinations of preemption points
or with different stack traces entirely arising from the same underlying cause,
course staff sent them all to the student,
including a disclaimer along the lines of ``some of these may indicate the same bug''.
%
The other 3 tests ({\tt priority-sema}, {\tt alarm-simultaneous}, and {\tt wait-twice}) found no bugs among the 4 groups.
One of the 4 groups had no bugs found among any of the 5 tests.

@@

% XXX: did I actually fix this in the annotation scripts, or did I just hackhackhack to suppress it?
% my email to kevin just says, "I suppressed the false positive bug from wait-simple in GROUPNAME and got a real bug next".
Finally, one false positive was found while testing {\tt wait-simple}.
Landslide mistakenly reported that {\tt free()} had reentered {\tt malloc()}
(normally a heap corruption bug)
due to a technical discrepancy between when Pebbles and Pintos kernels take and release their heap allocation lock
with regard to the rest of the {\tt malloc()} API.
After discarding the bug report and suppressing this false positive,
Landslide moved on to find the true deterministic use-after-free described above.

Overall, I consider all of the 7 bugs to be ``severe'':
they are all either correctness violations (priority mis-donation)
or stability issues (crash, deadlock, or data corruption).
In the upcoming survey section, although only one of these groups did the survey,
they enthusiastically reported finding the bug reports very helpful (\sect{\ref{sec:education-survey-response-data}}).
Despite the small sample size, I consider this
a positive result that Landslide can be a useful concurrency programming aid
even at universities with different class projects than CMU's 15-410.
As for 1 false positive among 8 total reports,
it is difficult to decide what rate of mistakes is acceptable \cite{racerx},
but the upcoming survey results (\sect{\ref{sec:education-eval-survey-falsepositives}})
suggest that students are generally at least patient enough
to ask for help and make progress when technical support is available.

@@

Figure~\ref{fig:photo-of-ze-studence} shows the distribution of project grades in CMU 15-410
between Fall 2013 and Spring 2018,
grouped by whether or not students used Landslide during P2.
%Landslide users are, of course, the experimental group.
The control group is further distinguished by whether Landslide was offered and the student delined,
or whether the study had not yet begun that semester.
Intuition suggests students in the former group would be more likely
to be struggling too much to have free time to volunteer in the first place,
so any comparison between them and Landslide users is vulnerable to selection bias.
The latter control group, in which students did not have a choice, mitigates such bias,
although itself may be vulnerable to other confounding factors such as grading criteria varying across semesters.
%
In addition to P2, I also show the subsequent P3 (kernel project) grades,
likewise broken down by who used Landslide previously during P2%
\footnote{The P3 distributions are slightly smaller than those from P2,
as some students dropped the class in between.
A few new project groups also tend to appear during P3 as partnerships break up or re-form,
but as these cannot be meaningfully classified as Landslide users or non-users,
I omit them here.
}.
Should Landslide use be correlated with higher grades on the later, more difficult concurrency project,
one explanation could be that Landslide helped students
internalize new concurrency programming and/or debugging skills
that would help them write better kernels even without Landslide's aid.

@@

{\bf Statistical significance.}
Table~\ref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess whether the differences are significant,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
testing if their samples are likely to arise from the same underlying, though unspecified, distribution%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between EDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}.
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these 4 comparsons like I do in the next section
because each tests a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses (such as ``more skilled students were more likely to sign up to begin with'').
%
Ultimately, only the same-semester comparisons were significant.
However, the distribution difference being smaller in P3 than in P2
% TODO: is there a good way to treat this more formally?
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias,
or else a similar difference should have been observed in the P3 distributions.
Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.

@@

@@

Based on the results from \sect{\ref{sec:education-eval-bugfinding}},
I selected 3 bugs to study in depth to ascertain
whether Landslide played an instrumental difference
in helping the students ultimately submit respecitvely correct implementations.
To avoid bias of picking too obscure and/or trivial bugs
that Landslide alone might find but even course staff would not expect students to solve,
I chose only bugs which had substantial penalties in the grading rubric
(guided, as well, by my own intuition as a former TA).
While checking for a given bug's presence by manual inspection,
I blinded myself to whether each group had been a Landslide user or not.
After unblinding, I then also classified students who used Landslide in general,
but whose usage snapshots showed they did not run the test case in question,
as non-users.
Table~\ref{tab:eval-common-bugs} presents the results.

@@

\item
{\bf Paraguay.}
Named after Ivan Jager, the Paraguayan 15-410 TA from 2004-2006 who originally discovered it,
this refers to a condition-variable bug in which a thread which sequentially sleeps on two different condition variables
can spuriously wake up early from the second {\tt cond\_wait()}.
The precise reasoning why many na\"ive student implementations are susceptible to this bug,
as well as the 3 major ways of fixing it,
are one of 15-410 staff's closely-guarded secrets (\sect{\ref{sec:410-secrecy}}).
Suffice it to say that the {\tt paraguay} test invokes a custom Pathos {\tt misbehave} mode
which biases thread scheduling towards the interleaving required to exhibit the bug
(Landslide, of course, replaces this {\tt misbehav}iour with model checking).
% TODO: i feel tihs sentence should go somewhere else, like after analysis, not in description
Hence, despite being a subtle concurrency bug,
the official course test suite is likely to expose it,
so comparing how many Landslide users and non-users submitted this bug in particular
speaks more to the impact of Landslide's preemption traces
in helping students diagnose a bug that a ``stress'' test could already find.
%
{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug.
%, and
%the official rubric penalty for it is
%{\tt :paraguay -2}.
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.

@@

On the other hand, because the {\tt thr\_exit()} bug typically manifests as a use-after-free,
the official tests must stress the thread library until the memory in question gets re-allocated
in just the right interleaving pattern to cause corruption that leads to a visible crash.
Landslide, on the other hand, detects this error immediately upon access (\sect{\ref{sec:landslide-valgrind-mode}}),
with no need for such complex corruption conditions.
Among the 11 Landslide users who submitted correct {\tt thr\_exit()}s in this regard,
it is difficult to say whether the heap checking alone, together with unit or stress tests,
would have been sufficient,
or whether heap checking and model checking were both necessary in concert to help them.
To isolate the impact of model checking,
I simulated the students having access to a stand-alone heap checker
by checking only the first thread interleaving with Landslide
% TODO - more specific section reference - quicksand det/nondet drpps eval
(similarly to \sect{\ref{sec:quicksand-eval}})
and assuming the students would find and fix any such ``deterministic'' use-after-free bugs before the deadline.
Re-classifying these into the ``correct'' group,
the 11-10-0-16 distribution becomes
% fisher.test(rbind(c(11,10),c(3,13)))$p.value -->  0.0475343
11-10-3-13,
with a new $p$ value of 0.048:
still positively correlated, but no longer significant under the multiplicity-corrected cutoff.

@@

\noindent
Overall, most of the issues reported as answers to the ``false positive'' survey question
were merely Landslide crashes or user interface confusion.
Those few truly erroneous bug reports (items 1, 2, and 7),
while certainly guilty of burdening % oopsative mood
students with worry over whether the problem is their own code or in Landslide itself,
are at least not too discouraging for two reasons.
Firstly, in all cases the students were able to recognize the false positive quickly enough to ask for help,
and I was able to deploy a fix and let them proceed before the project came due.
Secondly, each such problem, now having been fixed, will befall no future student again --
not to assert Landslide is completely bug-free now,
but it at least grows more and more stable with each passing semester.

@@

After the question asking ``Would you recommend a friend taking OS next semester to use Landslide?'',
I asked the students for open-form reasons why or why not.
As the former question's answers were exclusively positive (only 1 student even answering ``no opinion''),
this question's answers turned out to be mostly praise.
3 students declined to answer this question;
I reproduce the rest, paraphrasing for clarity and brevity,%
\footnote{Note that most students used the term ``race condition'' rather than ``concurrency bug'',
as taught in CMU and PSU lecture material;
I replaced these while paraphrasing in accordance with \sect{\ref{sec:glossary}}.}
here.
Answers from CMU:

@@

\subsection{Other universities}

@@

Human subjects research is inherently messy.
Each of the previous section's approaches to evaluating Landslide's educational value
was accompanied by some drawback which prevented it from being perfectly objective science,
but many of them presented tentatively positive results nonetheless.
Landslide helped many students find and fix many bugs (\sect{\ref{sec:education-eval-bugfinding}}),
but making a direct comparison to stress testing, the prior state of the art, is not straightforward.
%and not statistically significant,
Immediate improvement in students' project grades was observed,
(\sect{\ref{sec:education-eval-grades}}),
although statistical significance was lost when attempting to account for selection bias;
and impact on grades alone is a very narrow measure of pedagogical value anyway.
Landslide's debugging power was also found to be statistically significant
for the {\tt thr\_exit\_join} bug in particular.
Students responded overwhelmingly positively in the survey (\sect{\ref{sec:education-eval-survey}}),
although it is easy to imagine students being equally happy with a ``debugging tool'' that just tells them all the answers.

@@

I attribute the low participation rate of Pintos students to P2 students to two major factors:
one, not incentivizing the students to directly improve their grades
(instead offering only the vague promise of a ``learning experience'' debugging their code only after handin),
and two, not traveling to the university to introduce the research topic in an in-person lecture
(leaving the students potentially confused about what advantage, if any, was offered over stress testing).
Hypothetically, I could have achieved greater user study participation
either by offering extra credit to students
or by offering an autograder-like interface for students to receive bug reports before their deadlines instead of after
(either way requiring a more rigorous IRB review process).

>>>>

The (unpatched versions of) the first four tests are available at
\url{https://github.com/bblum/pintos} (a fork of the main Pintos basecode repository).
%\url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
The fifth test is available at
\url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.

@@

\begin{table}[h]
	\begin{center}
		% ugh this looks so stupid
		% but better than latex's output looking stupid
		\begin{tabular}{r|crcrc|r}
			& \multicolumn{5}{c|}{\bf Used Landslide?} & \\
			\bf Semester & & \bf yes & & \bf no & & \bf Total \\
			\hline
			CMU S'15	& &   8	& &  18	& &  26 \\
			CMU F'15	& &  22	& &   8	& &  30 \\
			CMU S'16	& &  18	& &  16	& &  34 \\
			CMU F'16	& &  12	& &   6	& &  18 \\
			CMU S'17	& &  19	& &  13	& &  32 \\
			CMU F'17	& &  14	& &   5	& &  19 \\
			CMU S'18	& &  10	& &   8	& &  18 \\
			\hline
			U. Chicago F'17	& &   4	& &  17	& &  21 \\ % 21 taken from number of submitted 'threads', idk abt userprog
			% TODO when you adjust this, adjust the % of participance down in the discussion para
			PSU S'18	& &  38	& &  98	& & 136 \\
			\hline
			\bf CMU total	& & 103	& &  74	& & 177 \\
			\bf Non-CMU total
					& &  42	& & 115	& & 157 \\
			\bf Total	& & 145	& & 189	& & 327 \\
		\end{tabular}
	\end{center}
	\caption{Participation rates across semesters,
	among students who submitted thread libraries (CMU, PSU) or kernels (U. Chicago).}
	\label{tab:photo-of-ze-studence}
\end{table}

@@

In the recruiting lectures for every semester after the first,
I included tallies of Landslide's bug-finding achievements to date
as an extra advertisement to entice students to participate.
These included each of the measures I counted by hand as described above,
as well as aggregate totals of how many groups participated,
how many groups found bugs,
how many found at least one concurrency bug,
and how many were able to fix and subsequently verify any or all thereof.
Among the concurrency bugs, a wide variety of types were found:
deadlocks, use-after-frees, segfaults, infinite loops, assertion failure, and unit test failure.
Table~\ref{tab:cmubugs} shows these statistics for each semester.
% may also be an overestimate if they only hacked around the bug, but that's less likely
Note that the tallies of bugs fixed may be undercounting
on account of the possibility some students may have truly fixed them but skipped the verification step thereafter.

\begin{table}[t]
	% TODO
	\begin{center}
		\small
	\begin{tabular}{r||r|r|r|r|r|r|r||r}
			& \bf S'15 & \bf F'15 & \bf S'16 & \bf F'16 & \bf S'17 & \bf F'17 & \bf S'18 & \bf Total \\
			\hline
			%			S15	F15	S16	F16	S17	F17	S18
			\bf Participating groups& 8	& 22	& 18	& 12	& 19	& 14	& 10	& 103	\\
			% nb not counting FPs
			Groups w/any bugs found	& 5	& 21	& 12	& 7	& 14	& 10	& 6	& 75	\\
			Groups w/any bugs fixed	& 4	& 18	& 9	& 6	& 9	& 7	& 5	& 58	\\
			Groups w/all ($\ne$0) bugs fixed
						& 2	& 11	& 3	& 3	& 7	& 5	& 4	& 35	\\
			\hline
			% the 2 for f15 were pretty messy; i had to go back and get em by hand by grepping for NOTE
			% but this semester also had a LOT of seemingly true race NOTE mx lock inf loop,
			% so i counted only the det ones, of which there were 2 (both also in "fixed" category)
			% otherwise 1 race recursive mx and 6 race mutex loops and 1 race cvar loop were left as races.

@@

This table also shows the number of false positives, as self-reported by Landslide
as described in \sect{\ref{sec:education-pebbles-instrumentation}}.
% nah where actually doing this, where making it hapen.
%which I started counting since S'16.
Please note that this approach %to counting false positives
is limited by Landslide's ability to classify them as ``suspicious'', even if not definitely bugs.
These represent technical obstacles that either
prevented Landslide from being able to analyze the synchronization in play
(e.g., recursive {\tt mutex\_lock()} invocations)
or were deemed too pedagogically important to allow the student to proceed without fixing
(e.g., busy spin-wait synchronization loops)%
\footnote{
Landslide was also configured to issue a ``bug report''
if {\tt MAGIC\_BREAK}, a Simics debugging trap, was ever invoked,
with specific instructions to remove it;
I consider these closer in spirit to compilation errors than bug reports and so did not count them
even among false positives.
}.
Other false positives arising from bugs in Landslide itself
required more individual effort to confirm;
to classify these I relied on students to help report them
during the Landslide clinics and survey,
and I report on them in \sect{\ref{sec:education-eval-survey-falsepositives}}.
Nevertheless,
apart from the spin-wait synchronization loops
(which can arise nondeterministically due to lock contention, but which Landslide already self-identifies),
I am aware of no cases of false positive {\em nondeterministic} bugs;
all unexpected false positives encountered to date were output on the first interleaving tested.
Though unscientific,
this provides some assurance that Table~\ref{tab:cmubugs}'s count of concurrency bugs is accurate,
even if the deterministic tally may overcount.

@@

In cases where Landslide emitted multiple preemption traces for the same bug,
whether manifested the same way through different combinations of preemption points
or with different stack traces entirely arising from the same underlying cause,
course staff sent them all to the student,
including a disclaimer along the lines of ``some of these may indicate the same bug''.
%
The other 3 tests, {\tt priority-sema}, {\tt alarm-simultaneous}, and {\tt wait-twice},
found no bugs among the 4 groups
({\tt wait-twice} being run only on the group that passed {\tt wait-simple}).
One of the 4 groups had no bugs found among any of the 5 tests.

@@

% XXX: did I actually fix this in the annotation scripts, or did I just hackhackhack to suppress it?
% my email to kevin just says, "I suppressed the false positive bug from wait-simple in GROUPNAME and got a real bug next".
Finally, one false positive was found while testing {\tt wait-simple}.
Landslide mistakenly reported that {\tt free()} had reentered {\tt malloc()}
due to a technical discrepancy between when Pebbles and Pintos kernels take and release their heap allocation lock
with regard to the rest of the {\tt malloc()} API.
This typically indicates a heap corruption bug, but in this case,
{\tt malloc()} was preempted after it released the heap lock, which was perfectly safe.
After discarding the bug report and suppressing this false positive,
Landslide moved on to find the true deterministic use-after-free described above.

Overall, I consider all of the 7 bugs to be ``severe'':
they are all either correctness violations (priority mis-donation)
or stability issues (crash, deadlock, or data corruption).
In the upcoming survey section, although only one of these groups did the survey,
they enthusiastically reported finding the bug reports very helpful (\sect{\ref{sec:education-survey-response-data}}).
Despite the small sample size, I consider this
a positive result that Landslide can be a useful concurrency programming aid
even at universities with different class projects than CMU's 15-410.
As for 1 false positive among 8 total reports,
it is difficult to decide what rate of mistakes is acceptable when the user's patience is at stake \cite{racerx},
but the upcoming survey results (\sect{\ref{sec:education-eval-survey-falsepositives}})
suggest that students are generally at least willing
to ask for help and make progress when technical support is available.

@@

Figure~\ref{fig:photo-of-ze-studence} shows the distribution of project grades in CMU 15-410
between Fall 2013 and Spring 2018,
grouped by whether or not students used Landslide during P2.
%Landslide users are, of course, the experimental group.
The control group is further distinguished by whether Landslide was offered and the student declined,
or whether the study had not yet begun that semester.
Intuition suggests students in the former group would be more likely
to be struggling too much to have free time to volunteer in the first place,
so any comparison between them and Landslide users is vulnerable to selection bias.
The latter control group, in which students did not have a choice, mitigates such bias,
although itself may be vulnerable to other confounding factors such as grading criteria varying across semesters.
%
In addition to P2, I also show the subsequent P3 (kernel project) grades,
likewise broken down by who used Landslide previously during P2%
\footnote{The P3 distributions are slightly smaller than those from P2,
as some students dropped the class in between.
Some new project groups also tend to appear during P3 as students either solo or find new partners,
but as these cannot be meaningfully classified as Landslide users or non-users,
I omit them here.
}.
Should Landslide use be correlated with higher grades on the later, more difficult concurrency project,
one explanation could be that Landslide helped students
internalize new concurrency programming and/or debugging skills
that would help them write better kernels even without Landslide's aid.

@@

{\bf Statistical significance.}
Table~\ref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess whether the differences are significant,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
testing if their samples are likely to arise from the same underlying, though unspecified, distribution%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between EDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}.
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these 4 comparsons like I do in the next section
because each tests a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses,
such as ``more skilled students were more likely to sign up to begin with''.
%
Ultimately, only the same-semester comparisons were significant.
However, the distribution difference being smaller in P3 than in P2
% TODO: is there a good way to treat this more formally?
% TODO: run a combined A-D test for each of the two control pairs
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias,
or else a similar difference should have been observed in the P3 distributions.
Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.

@@

% combined test
% > ad.test.combined(list(c(42,50,51,55,55,55,56,57,57,58,62,63,65,65,65,66,66,67,67,67,67,68,68,68,69,69,69,70,70,70,70,71,71,71,71,71,72,73,73,73,73,73,74,74,74,74,74,75,75,75,75,75,76,76,76,76,77,77,77,77,78,78,78,78,79,79,79,79,79,79,80,80,80,80,81,81,81,82,82,82,82,83,84,84,84,84,85,85,86,86,86,87,87,87,88,89,90,91,92,93,93,93,95),c(43,51,56,58,59,59,60,60,60,60,61,61,61,61,61,62,63,65,65,67,68,69,69,69,69,69,70,70,71,71,72,72,72,73,73,73,73,73,74,74,74,74,75,75,75,75,76,76,77,77,77,77,78,78,78,78,79,79,79,80,80,80,80,80,80,81,81,82,82,82,82,83,83,83,83,85,85,85,86,86,86,88,88,90,92,93)), list(c(65,68,72,72,73,75,75,75,75,75,76,76,76,76,77,77,77,78,78,78,78,78,78,79,79,80,80,80,80,80,80,80,81,82,82,82,82,82,82,83,83,83,83,83,83,84,84,84,84,85,85,85,85,85,85,85,85,85,85,86,86,86,86,86,86,86,86,86,87,87,87,88,88,88,88,88,89,89,89,89,90,90,90,90,90,91,91,91,91,92,92,93,93,93,94,95,95,95),c(60,62,63,64,65,68,69,72,72,74,75,75,75,75,76,76,77,77,77,77,78,78,78,78,79,80,80,80,81,81,81,81,81,81,81,81,81,82,82,82,82,82,83,83,83,84,84,84,84,84,84,84,85,86,86,87,87,88,88,88,88,88,88,88,89,89,89,89,89,90,90,90,91,92,92,92,93,94,95)))
%            AD.comb T.comb  asympt. P-value
% version 1:   2.176 0.1655           0.3324
% version 2:   2.219 0.2055           0.3191
%
% > ad.test.combined(list(c(42,50,51,55,55,55,56,57,57,58,62,63,65,65,65,66,66,67,67,67,67,68,68,68,69,69,69,70,70,70,70,71,71,71,71,71,72,73,73,73,73,73,74,74,74,74,74,75,75,75,75,75,76,76,76,76,77,77,77,77,78,78,78,78,79,79,79,79,79,79,80,80,80,80,81,81,81,82,82,82,82,83,84,84,84,84,85,85,86,86,86,87,87,87,88,89,90,91,92,93,93,93,95),c(38,42,47,50,52,53,55,55,56,56,57,57,58,58,58,59,60,60,61,62,62,63,65,66,67,68,68,68,69,70,70,70,71,71,71,71,72,72,72,73,73,73,74,74,76,76,76,77,77,78,78,78,78,78,79,79,79,80,82,82,83,83,83,84,84,84,85,86,86,88,90,95,96,96)),list(c(65,68,72,72,73,75,75,75,75,75,76,76,76,76,77,77,77,78,78,78,78,78,78,79,79,80,80,80,80,80,80,80,81,82,82,82,82,82,82,83,83,83,83,83,83,84,84,84,84,85,85,85,85,85,85,85,85,85,85,86,86,86,86,86,86,86,86,86,87,87,87,88,88,88,88,88,89,89,89,89,90,90,90,90,90,91,91,91,91,92,92,93,93,93,94,95,95,95),c(45,58,66,70,72,72,72,72,74,74,74,75,75,75,77,77,77,78,78,78,79,79,79,79,79,79,80,80,81,81,81,81,81,82,82,83,83,83,83,83,84,84,85,85,86,86,86,86,86,86,87,87,89,89,89,89,90,91,91,91,93,93,93)))
%            AD.comb T.comb  asympt. P-value
% version 1:   5.256  3.058          0.01503
% version 2:   5.560  3.343          0.01109

@@

Based on the results from \sect{\ref{sec:education-eval-bugfinding}},
I selected 4 bugs to study in depth to ascertain
whether Landslide played an instrumental difference
in helping the students ultimately submit respectively correct implementations.
To avoid bias of picking too obscure and/or trivial bugs
that Landslide alone might find but even course staff would not expect students to solve,
I chose only bugs which had substantial penalties in the grading rubric
(guided, as well, by my own intuition as a former TA).
While checking for a given bug's presence by manual inspection,
I blinded myself to whether each group had been a Landslide user or not.
After unblinding, I then re-classified students who used Landslide in general,
but whose usage snapshots showed they did not run the test case in question,
as non-users.
Table~\ref{tab:eval-common-bugs} presents the results.

@@

\item
{\bf Paraguay.}
Named after Ivan Jager, the Paraguayan 15-410 TA from 2004-2006 who originally discovered it,
this refers to a condition-variable bug in which a thread which sequentially sleeps on two different condition variables
can spuriously wake up early from the second {\tt cond\_wait()}.
The precise reasoning why many na\"ive student implementations are susceptible to this bug,
as well as the 3 major ways of fixing it,
are one of 15-410 staff's closely-guarded secrets (\sect{\ref{sec:410-secrecy}}).
Suffice it to say that the {\tt paraguay} test invokes a custom Pathos {\tt misbehave} mode
which biases thread scheduling towards the interleaving required to exhibit the bug
(Landslide, of course, replaces this {\tt misbehav}iour with model checking).
% TODO: i feel tihs sentence should go somewhere else, like after analysis, not in description
Hence, despite being a subtle concurrency bug,
the official course test suite is likely to expose it,
so comparing how many Landslide users and non-users submitted this bug in particular
would speak more to the impact of Landslide's preemption traces
in helping to diagnose a bug that a ``stress'' test could already find.
%
{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug.
%, and
%the official rubric penalty for it is
%{\tt :paraguay -2}.
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.

@@

On the other hand, because the {\tt thr\_exit()} bug typically manifests as a use-after-free,
the official tests must stress the thread library until the memory in question gets re-allocated
in just the right interleaving pattern to cause corruption that leads to a visible crash.
Landslide, meanwhile, detects this error immediately upon access (\sect{\ref{sec:landslide-valgrind-mode}}),
with no need for complex corruption conditions.
Among the 11 Landslide users who submitted correct {\tt thr\_exit()}s in this regard,
it is difficult to say whether the heap checking alone, together with unit or stress tests,
would have been sufficient,
or whether heap checking and model checking were both necessary in concert to help them.
To isolate the impact of model checking,
I simulated the students having access to a stand-alone heap checker
by checking only the first thread interleaving with Landslide
% TODO - more specific section reference - quicksand det/nondet drpps eval
(similarly to \sect{\ref{sec:quicksand-eval}})
and assuming the students would find and fix any such ``deterministic'' use-after-free bugs before the deadline.
Re-classifying these into the ``correct'' group,
the 11-10-0-16 distribution becomes
% fisher.test(rbind(c(11,10),c(3,13)))$p.value -->  0.0475343
11-10-3-13,
with a new $p$ value of 0.048:
still positively correlated, but no longer significant under the multiplicity-corrected cutoff.

@@

\noindent
Overall, most of the issues reported as answers to the ``false positive'' survey question
were merely Landslide crashes or user interface confusion.
Those few truly erroneous bug reports (items 1, 2, and 7),
while certainly guilty of burdening % oopsative mood
students with worry over whether the problem is their own code or in Landslide itself,
are at least not too discouraging for two reasons.
Firstly, in all cases the students were able to recognize the false positive quickly enough to ask for help,
and I was able to deploy a fix and let them proceed before the project came due.
Secondly, each such problem, now having been fixed, will befall no future student again --
not to assert Landslide is completely bug-free now,
but at least that it grows more and more stable with each passing semester.

@@

After the question asking ``Would you recommend a friend taking OS next semester to use Landslide?'',
I asked the students for open-form reasons why or why not.
As the former question's answers were exclusively positive (only 1 student even answering ``no opinion''),
this question's answers turned out to be mostly praise.
3 students declined to answer this question;
I reproduce the rest here, paraphrasing for clarity and brevity%
\footnote{Note that most students used the term ``race condition'' rather than ``concurrency bug'',
as taught in CMU and PSU lecture material;
I replaced these while paraphrasing in accordance with \sect{\ref{sec:glossary}}.}.
Answers from CMU:

@@

\subsubsection{Other universities}

@@

Human subjects research is inherently messy.
Each of the previous section's approaches to evaluating Landslide's educational value
was accompanied by some drawback which prevented it from being perfectly objective science,
but many of them presented tentatively positive results nonetheless.
Landslide helped many students find and fix many bugs (\sect{\ref{sec:education-eval-bugfinding}}),
but making a direct comparison to stress testing, the prior state of the art, is not straightforward.
%and not statistically significant,
Immediate improvement in students' project grades was observed
(\sect{\ref{sec:education-eval-grades}}),
although statistical significance was lost when attempting to account for selection bias;
and impact on grades alone is a very narrow measure of pedagogical value anyway.
Landslide's debugging power was also found to be statistically significant
for the {\tt thr\_exit\_join} bug in particular.
Students responded overwhelmingly positively in the survey (\sect{\ref{sec:education-eval-survey}}),
although it is easy to imagine students being equally happy with a ``debugging tool'' that just tells them all the answers.

@@

I attribute the comparatively low participation rate of Pintos students to two major factors:
one, not incentivizing the students to directly improve their grades
(instead offering only the vague promise of a ``learning experience'' debugging their code only after handin),
and two, not traveling to the university to introduce the research topic in an in-person lecture
(leaving the students potentially confused about what advantage, if any, was offered over stress testing).
Hypothetically, I could have achieved greater user study participation
either by offering extra credit to students
or by offering an autograder-like interface for students to receive bug reports before their deadlines instead of after
(either way requiring a more rigorous IRB review process).
